#######################
## Problem Set 6     ##
#######################

**Due Date**: February 20, 2026  
**Submission**: [https://canvas.northwestern.edu/courses/245562/assignments/1687751](https://canvas.northwestern.edu/courses/245562/assignments/1687751)


---

####################
## Problem 1      ##
####################


**1a.**
Using the democracy and internet access data from the lecture:

```{r problem1a, eval=FALSE}
# Load data and run models
library(rqog)
qogts <- read_qog(which_data = "standard", data_type = "time-series")

# Model 1: Basic model with homoskedastic errors
model1 <- lm(vdem_libdem ~ wdi_broadb + I(log(wdi_gdpcappppcon2017)), 
             data = qogts)

# Model 2: Robust standard errors
```

**Questions:**
1. Extract and compare the standard errors. Which are larger?
2. Calculate the t-statistics manually (i.e., using a calculator or direct math in R and not just summary): $t = \frac{\hat{\beta}}{SE(\hat{\beta})}$
3. Compute p-values for both sets of standard errors
4. At α = 0.05, would you reject the null hypothesis for each coefficient in both models?


**1b.**
The lecture notes that with small samples and normal errors, we use the t-distribution.

1. What is the degrees of freedom for the t-distribution in your model?
2. How do t-critical values compare to z-critical values for your sample size? How much difference does it make to choose the t versus normal distribution at this sample size?


---

####################
## Problem 2      ##
####################

**2a.**
1. In your own words, explain the multiple comparisons problem
2. Define:
   - Family-Wise Error Rate (FWER)
   - False Discovery Rate (FDR)

**2b.**
Simulate the multiple comparisons problem:

```{r problem2b, eval=FALSE}
set.seed(123)
n_tests <- 20
n_obs <- 100
alpha <- 0.05

# Create matrix of 20 independent tests (all null true)
p_values <- matrix(NA, nrow = 1000, ncol = n_tests)

for (i in 1:1000) {
  for (j in 1:n_tests) {
    # Generate independent data
    x <- rnorm(n_obs)
    y <- rnorm(n_obs)  # No relationship
    # Run regression and extract p-value for slope
    p_values[i, j] <- summary(lm(y ~ x))$coefficients[2, 4]
  }
}

# Analyze results
false_positives <- rowSums(p_values < alpha)
mean_fp <- mean(false_positives)
prop_at_least_one <- mean(false_positives > 0)
```

**Questions:**
1. What proportion of simulations have at least one false positive?
2. What is the average number of false positives per simulation?
3. Create a histogram of the number of false positives across simulations.

**2c.**
Now simulate correlated tests:

```{r problem3c, eval=FALSE}
# Create correlated predictors
set.seed(123)
n_tests <- 20
n_obs <- 100

# Generate correlated X matrix
library(MASS)
mu <- rep(0, n_tests)
Sigma <- diag(n_tests)
for (i in 1:n_tests) {
  for (j in 1:n_tests) {
    Sigma[i, j] <- 0.7^abs(i-j)  # AR(1) correlation
  }
}

p_values_cor <- matrix(NA, nrow = 1000, ncol = n_tests)

for (i in 1:1000) {
  # Generate correlated predictors
  X <- mvrnorm(n_obs, mu, Sigma)
  # Generate Y with no relationship to any X
  y <- rnorm(n_obs)
  
  # Run all regressions
  for (j in 1:n_tests) {
    p_values_cor[i, j] <- summary(lm(y ~ X[, j]))$coefficients[2, 4]
  }
}

# Compare with independent case
false_positives_cor <- rowSums(p_values_cor < alpha)
prop_cor <- mean(false_positives_cor > 0)
```

**Questions:**
1. How does correlation affect the multiple comparisons problem?

---

####################
## Problem 3      ##
####################

**3a.**
Find a published political science article that reports multiple hypothesis tests (e.g., a table with many coefficients and stars).

1. Count how many hypothesis tests are reported
2. Calculate the expected number of false positives if all nulls were true
3. Would multiple testing corrections change their conclusions?

**3b.**
You're designing a study to test 15 different hypotheses about voter behavior.

1. How would you adjust your analysis plan to account for multiple comparisons?
2. Would you use FWER or FDR control? Justify your choice.
3. How would sample size affect your decision?
4. What would you report in the methods section about multiple testing?

